Evolution of precopulatory and postcopulatory inbreeding behaviour

### 2016-08-31 ms_pre_post.Rmd
Revise Model analysis. Outline introductory paragraphs

Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate traits including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use *in silico* experiments to test the hypotheses that evolution of post-copulatory inbreeding avoidance facilitates evolution of polyandry, and that evolution of pre-copulatory versus post-copulatory inbreeding avoidance can be predicted by the relative costs of each, and the relative cost of polyandry. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and different magnitudes of inbreeding depression in offspring viability.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across a range of costs and inbreeding depression slopes using two types of simulations. First, we consider eight possible cost parameter combinations in which costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal either $0$ or $0.02$ [@Duthie observed $0.02$ to be strong cost of tendency for polyandry]. Second, we fix $c_{P}$ to values of $0$, $0.01$, or $0.02$, then randomly select $c_{M}$ and $c_{F}$ values between $0-0.05$ for simulation initial starting conditions. The objective of the first type of simulation is to test how mean allele values change over generations in response to different combinations of costs and inbreeding depression slopes. The objective the second type of simulation is to predict selection for tendency for polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy, across a wider range of parameter combinations, and to therefore test under what conditions each is predicted to evolve given different combinations of costs and inbreeding depression slopes.

In the first type of simulation, we record dynamics of mean values of $P_{a}$, $M_{a}$, $F_{a}$ over generations for every possible combination in which the cost associated with each of the three phenotypes ($c_{P}$, $c_{M}$, and $c_{F}$) is either $0$ or $0.02$, and for four slopes of inbreeding depression on offspring viability $\beta = {0, 1, 2, 3}$. Each possible cost combination is replicated ten times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare).

In the second type of simulation, we record and present mean values of $P_{a}$, $M_{a}$, $F_{a}$ only from the terminal generation of simulations, and for random costs of inbreeding mating strategy $c_{M} \in [0, 0.05]$ and inbreeding fertilisation strategy $c_{F} \in [0, 0.05]$, but fixed costs of tendency for polyandry $c_{P} = {0, 0.01, 0.02}$. We consider four values of $\beta = {0, 1, 2, 3}$, and we simulate 100 replicates for all combinations of $c_{P}$ and $\beta$. For different values of $\beta$ given the same $c_{P}$ value, we use the same randomly selected values of $c_{M}$ and $c_{F}$ to better infer how terminal mean allele values are affected by different magnitudes of inbreeding depression. We do not use the same randomly selected values of $c_{M}$ and $c_{F}$ across different values of $c_{P}$ because $c_{P}$ values were fixed *a priori* only to facilitate model inference. The objective of randomising costs was to sample from a broad range of parameter space, and this range is increased by selecting different $c_{M}$ and $c_{F}$ values for each $c_{P}$.

For each replicate simulation with a unique set of costs, we classify the evolutionary outcome of inbreeding stragies in the simulation in one of four ways: (1) Weak selection for inbreding mating strategy and inbreeding fertilisation strategy, (2) weak selection for inbreeding mating strategy and strong selection for inbreeding fertilisation strategy, (3) strong selection for inbreeding mating strategy and weak selection for inbreeding fertilisation strategy, and (4) strong selection for inbreeding mating strategy and inbreeding fertilisation strategy. We additionally classify the evolutionary outcome of tendency for polyandry in the simulation as either weak selection for polyandry or strong selection for polyandry. @Duthie showed that selection for inbreeding preference (positive values of $M_{a}$) is predicted when inbreeding depression on offspring viability is $\beta=0$, but selection for inbreeding avoidance (negative values of $M_{a}$) is predicted when $\beta \geq 1$. When referring to strong selection for inbreeding mating strategy and inbreeding fertilisation strategy, we therefore refer specifically to inbreeding preference when $\beta=0$ and inbreeding avoidance when $\beta \geq 1$. Strong selection is inferred when mean allele values exceed $1$ (when $\beta = 0$) or are less than $-1$ (when $\beta \geq 1$), otherwise selection is inferred to be weak. Exploratory simulations suggest that cutoff values of $1$ and $-1$ are conservative for inferring when selection is strong, and other cutoff values do not change the general pattern of evolutionary outcomes (1-4) across cost parameter values.

For all replicates in both the first and second simulation types, we set the maximum number of generations to 5000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation such that females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mating encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

>>>>

<!--- TODO: GENERAL PARAGRAPH INTRODUCING THE SCOPE AND KEY QUESTION --->

<!--- TODO: PARAGRAGH ON EVOLUTION OF POLYANDRY --->

<!--- TODO: PARAGRAGH ON EVOLUTION OF PRE-COPULATORY INBREEDING AVOIDANCE --->

<!--- TODO: PARAGRAGH ON EVOLUTION OF POST-COPULATORY INBREEDING AVOIDANCE --->

<!--- TODO: PARAGRAGH ON IDEA THAT POSTCOP SELECTION COULD DRIVE SELECTION FOR POLYANDRY --->

<!--- TODO: PARAGRAGH ON IDEA THAT RELATIVE COSTS WILL DETERMINE PRE VS POST --->

<!--- TODO: PARAGRAGH ON COSTS OF POLYANDRY, MATING, AND FERTILISATION STRATEGIES --->

Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate traits including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use *in silico* experiments to test the hypotheses that, given inbreeding depression in offspring viability, evolution of post-copulatory inbreeding avoidance causes selection for evolution of polyandry, and that pre-copulatory but not post-copulatory inbreeding avoidance will evolve when pre-copulatory inbreeding avoidance is less costly than the combined cost of post-copulatory inbreeding avoidance and polyandry, while post-copulatory but not pre-copulatory inbreeding avoidance will evolve when pre-copulatory inbreeding avoidance is more costly than the combined costs of post-copulatory inbreeding avoidance and polyandry. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and different magnitudes of inbreeding depression in offspring viability.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), inbreeding mating strategy ($M_{a}$), and inbreeding fertilisation strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different combinations of inbreeding depression slopes and costs to test our two focal hypotheses. The combinations of costs and inbreeding depression slopes that we consider include all eight possible permutations of values of costs of tendency for polyandry ($c_{P}$), inbreeding mating strategy ($c_{M}$), and inbreeding fertilisation strategy ($c_{F}$) equal to either $0$ or $0.02$ [@Duthie observed $0.02$ to be strong cost of tendency for polyandry], and for values of $\beta = \{1, 3\}$ ($16$ total combinations). In all of these simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each of the 16 possible cost combinations is replicated $40$ times, and grand mean values and standard deviation of means is calculated in each generation across replicate simulations. These analyses allowed us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on inbreeding fertilisation strategy might be weak if polyandry is rare).

To test our hypothesis that selection for evolution of polyandry is driven by female's ability to bias fertilisation toward inbreeding avoidance through post-copulatory mechanisms, we compare simulations in which all three phenotypes including tendency for polyandry ($P_{p}$), inbreeding mating strategy ($M_{p}$), and inbreeding fertilisation strategy ($F_{p}$) evolve, with an identical set of simulations in which $F_{p}$ cannot evolve. In simulations without evolving $F_{p}$, we therefore include all 16 combinations of $c_{P}=\{0,0.02\}$, $c_{M}=\{0,0.02\}$, $c_{F}=\{0,0.02\}$, and $\beta = \{1,3\}$ replicated 40 times each. To stop $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes resulting from $F_{a}$ cause random fertilisation with respect to kinship. Hence, while mutation and drift can still occur for $F_{a}$ alleles, these alleles have no phenotypic effect and polyandry cannot cause polyandrous females to benefit through an increased ability to bias fertilisation toward inbreeding avoidance. By comparing evolution of mean $P_{p}$ values over generations when $F_{p}$ also evolves versus when $F_{p}=0$ (random fertilisation), we test whether or not evolution of inbreeding fertilisation strategy increases selection for evolution of polyandry.

To test hypotheses regarding evolution under different relative costs of phenotypes, we consider all combinations of costs and inbreeding depression parameter values using only simulations in which all three traits evolve. To test our hypothesis that inbreeding mating strategy will evolve toward inbreeding avoidance ($M_{p}<0$) when mating costs are low relative to the costs of tendency for polyandry and inbreeding fertilisation strategy, but that selection for inbreeding fertilisation strategy will then be negligible ($F_{p} \approx 0$), we quantify evolution of $M_{p}$ and $F_{p}$ over generations in simulations where $c_{M}=0$ but $c_{P}$, $c_{F}$, or both equals $0.02$. We compare evolution in these three cost parameter value combinations with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $M_{p}$ but not $F_{p}$ will evolve to more negative values when $c_{M}$ is relatively low than when all cost values are uniform.

To test the complementary hypothesis that inbreeding fertilisation strategy will evolve toward inbreeding avoidance ($F_{p}<0$) given low costs of polyandry and inbreeding fertilisation strategy relative to the cost of inbreeding mating strategy, but that selection on inbreeding mating strategy will be negligible ($M_{p} \approx 0$), we qualtify evolution of $M_{p}$ and $F_{p}$ over generations when $c_{F}=0$, $c_{P}=0$, and $c_{M}=0.02$. We compare evolution for this combination of cost parameter values with simulations in which all costs equal either $0$ or $0.02$. If our hypothesis is supported, then $F_{p}$ but not $M_{p}$ will evolve to more negative values when $c_{F}$ is relative low than when all cost values are uniform.

For all replicates, we set the maximum number of generations to 10000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes. For all replicates, we set the number of immigrants per generation to $\rho=5$, which produced a range of kin and non-kin in each generation such that females could express inbreeding strategies. Females always produced $n=8$ offspring, which was sufficient to keep populations consistently near carrying capacities and avoid population extinction. Carrying capacities for females ($K_{f}$) and males ($K_{m}$) were both set to 100 because previous modelling has shown that populations of this size are small enough that mating encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).
